---
title: ""
bibliography:
csl: microbiome.csl
output: 
  pdf_document:
    keep_tex: true
fontsize: 12pt
geometry: margin=1.0in
---

# Differences in the Stool Microbiome Before and After Colorectal Cancer Treatment

\vspace{10mm}

\begin{center}
Marc A Sze, Nielson T Baxter, Mack T Ruffin IV, Mary AM Rogers, and Patrick D Schloss

\vspace{10mm}

Supplemental
\end{center}


\newpage

```{r knitr_settings, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

source('../code/functions.R')
source('../code/graphFunctions.R')

loadLibs(c("dplyr", "tidyr", "ggplot2", "reshape2", "gridExtra", "scales", "wesanderson", "knitr", "rmarkdown"))

```



**Table S1: P-values of Differences between Initial and Follow Up Samples for Richness, Shannon Diversity, and Evenness.**

```{r supplemental_table1, echo=FALSE, message=FALSE, warning=FALSE}

alpha_table <- read.csv("../results/tables/alpha_table_summary.csv", header = T, stringsAsFactors = F)
supplemental_table1 <- separate(alpha_table, X, c("comparison", "alpha_metric"), sep="_") 

colnames(supplemental_table1) <- c("Comparison", "Alpha Metric", "P-value", "BH Adjusted P-value")

kable(supplemental_table1, format = "markdown", align = 'c', padding = 2)
```


\newpage


**Table S2: Summary of OTUs in both Lesion and Initial/Follow Up Models**

```{r supplemental_table4, echo=FALSE, message=FALSE, warning=FALSE}

common_vars_summary <- read.csv('../results/tables/pvalue_IF_lesion_common_imp_vars.csv', header = T, row.names = 1, stringsAsFactors = F)

st2 <- select(common_vars_summary, otu) %>% 
  mutate(Tax = gsub("_unclassified", "", common_vars_summary$lowest_ID)) %>% 
  mutate(Pvalue = common_vars_summary$Pvalue) %>% 
  mutate(BH_corr = common_vars_summary$BH_corr)

colnames(st2) <- c("OTU", "Lowest Taxonomic ID", "P-value", "BH Corrected P-value")

kable(st2, format = "markdown", align = 'c', padding = 2)

```


\newpage


